For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/universaliit.tex
make uppercase greek letters italic

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $\mathsf{S}$.
\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\Gamma$ and a
    semantic context $\model{\Gamma}$ of the model $M$, we have
    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires weakening at the
middle of a context. We consider pairs of untyped contexts, which should be
thought of as a splitting of a context at some position. The full context is
recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
$\length{\Gamma}$ denotes the length of a the context $\Gamma$.

@@

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
  and $\Delta\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\Gamma$ as
  follows, where $\keep^{\length{\Gamma}}$ is $\keep$ iterated
  $\length{\Gamma}$ times:
\begin{alignat*}{5}
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \Gamma.\keep^{\length{\Gamma}} \epsilonu
  \end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

\subsubsection{Weakening preserves typing}
\label{ss:weaken-typing}
We prove by mutual induction that typing judgments are stable under weakening,
for contexts, types, terms, and substitutions.
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash) \ra
  \Gamma \extu A \merge \wkO\, \Delta \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, B  \\
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash t\in B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, t \in \wk_{\length{\Delta}}\,B  \\
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \vdash \sigma
  \Rightarrow \Delta) \ra
  \Gamma \extu A \vdash \wkO\, \sigma \Rightarrow \Delta
  \end{alignat*}
\subsubsection{Substitution preserves typing}
  We show that judgments are stable under substitution.
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash t \in A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\Delta \vdash x \invar A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\Gamma \vdash) \ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  (\Delta \vdash \tau \Rightarrow E) \ra
  \Gamma \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  % (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  % \Gamma \vdash A [ \sigma ]
  % \\
  \end{alignat*}
\subsubsection{Laws for identity substitutions}
We show category and functor laws involving identity substitution for
well-formed types, terms and substitutions.
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\Gamma \vdash A)\ra A [ \idu\,\Gamma]  \\
  & [\idu] && : (\Gamma \vdash x \invar A)\ra x [ \idu\,\Gamma] = V x \\
  & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma \\
  & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
  \end{alignat*}
Finally, we show that the identity substitution itself is well-typed:
  \begin{alignat*}{5}
  & \idw && : (\Gamma \vdash)\ra \Gamma \vdash \idu\,\Gamma \Rightarrow \Gamma
  \end{alignat*}

\subsubsection{Proof irrelevance and unicity of typing}
\label{ss:uniq-typing-types}
A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
  \[
    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
  \]
We show that each of the typing judgments is unique in the
following sense:
\begin{alignat*}{5}
 & \isp{\Conw} && : \isaprop\, (\Gamma\vdash) \\
 & \isp{\Tyw} && : \isaprop\, (\Gamma\vdash A) \\
 & \isp{\wf\Var} && : \isaprop\, (\Gamma\vdash x \invar A) \\
 & \isp{\Tmw} && : \isaprop\, (\Gamma\vdash t \in A) \\
 & \isp{\wf\Sub} && : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
\end{alignat*}
We also show unicity of typing:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\Gamma \vdash t \in A) \ra
  (\Gamma \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\Gamma \vdash x \invar A) \ra
  (\Gamma \vdash x \invar B) \ra A = B
  \end{alignat*}
Let us consider for instance the application constructor $\appw$: for a codomain
type $B$ it yields an overall type $C=B[\mathsf{0} := u]$ for an application. Even
if $C$ is known a priori, there may be another $B$ for which $B[\mathsf{0} := u] = C$,
possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$
of $t$.

@@

 \subsubsection{Syntactic model}
 The syntactic model is obtained by packing the untyped syntax with the typing
 judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\Gamma \Gamma\vdash
 \\
 & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
 \\
 & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
 \\
 & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
\end{alignat*}
The other fields are given straightforwardly.  Regarding the equations, it is
enough to prove them only for the untyped syntactic part: as we argued in
Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are
automatically equal.

@@

Given a model $M$ of the QIIT, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\Gamma\sim \model{\Gamma}$ equivalent to $\rec\,\,\Gamma=\model{\Gamma}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

>>>>

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $\mathsf{S}$.
\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\rec\,\,\GG
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
typing. Stating a typing rule for this operation requires weakening at the
middle of a context. We consider pairs of untyped contexts, which should be
thought of as a splitting of a context at some position. The full context is
recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\GG}$ of a
% context $\GG$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \GG \merge \cdot && :=  \GG \\
  & \GG \merge (\GD \extu A) && :=  (\GG \merge \GD)\extu A
\end{alignat*}
We think of the second context as a telescope over the first one. We also define
weakening for telescopes, which will be used to give typing rules for telescopes
in Section \ref{sec:typing_judgments}:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\GD\,\extu A) && := \wkO\,\GD\,\extu \wk_{\length{\GD}}\,A
  \end{alignat*}
$\length{\GG}$ denotes the length of a the context $\GG$.

@@

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\GG$ and
  $\GD$, then $\keep\,\sigma$ is a substitution between contexts $\GG\ext A[\sigma]$
  and $\GD\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\GG$ as
  follows, where $\keep^{\length{\GG}}$ is $\keep$ iterated
  $\length{\GG}$ times:
\begin{alignat*}{5}
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \GG.\keep^{\length{\GG}} \epsilonu
  \end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

\subsubsection{Weakening preserves typing}
\label{ss:weaken-typing}
We prove by mutual induction that typing judgments are stable under weakening,
for contexts, types, terms, and substitutions.
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash) \ra
  \GG \extu A \merge \wkO\, \GD \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, B  \\
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash t\in B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, t \in \wk_{\length{\GD}}\,B  \\
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \vdash \sigma
  \Rightarrow \GD) \ra
  \GG \extu A \vdash \wkO\, \sigma \Rightarrow \GD
  \end{alignat*}
\subsubsection{Substitution preserves typing}
  We show that judgments are stable under substitution.
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash t \in A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\GD \vdash x \invar A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\GG \vdash) \ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  (\GD \vdash \tau \Rightarrow E) \ra
  \GG \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  % (\GG \vdash \sigma \Rightarrow \GD) \ra
  % \GG \vdash A [ \sigma ]
  % \\
  \end{alignat*}
\subsubsection{Laws for identity substitutions}
We show category and functor laws involving identity substitution for
well-formed types, terms and substitutions.
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG]  \\
  & [\idu] && : (\GG \vdash x \invar A)\ra x [ \idu\,\GG] = V x \\
  & [\idu] && : (\GG \vdash t \in A)\ra t [ \idu\,\GG] = t \\
  & \idru && : (\GG \vdash \sigma \Rightarrow\GD)\ra \sigma \circ \idu\,\GG = \sigma \\
  & \idlu && : (\GG \vdash \sigma \Rightarrow\GD)\ra \idu\,\GD\circ \sigma = \sigma
  \end{alignat*}
Finally, we show that the identity substitution itself is well-typed:
  \begin{alignat*}{5}
  & \idw && : (\GG \vdash)\ra \GG \vdash \idu\,\GG \Rightarrow \GG
  \end{alignat*}

\subsubsection{Proof irrelevance and unicity of typing}
\label{ss:uniq-typing-types}
A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
  \[
    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
  \]
We show that each of the typing judgments is unique in the
following sense:
\begin{alignat*}{5}
 & \isp{\Conw} && : \isaprop\, (\GG\vdash) \\
 & \isp{\Tyw} && : \isaprop\, (\GG\vdash A) \\
 & \isp{\wf\Var} && : \isaprop\, (\GG\vdash x \invar A) \\
 & \isp{\Tmw} && : \isaprop\, (\GG\vdash t \in A) \\
 & \isp{\wf\Sub} && : \isaprop\, (\GG\vdash \sigma \Rightarrow \GD)
\end{alignat*}
We also show unicity of typing:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\GG \vdash t \in A) \ra
  (\GG \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\GG \vdash x \invar A) \ra
  (\GG \vdash x \invar B) \ra A = B
  \end{alignat*}
Let us consider for instance the application constructor $\appw$: for a codomain
type $B$ it yields an overall type $C=B[\mathsf{0} := u]$ for an application. Even
if $C$ is known a priori, there may be another $B$ for which $B[\mathsf{0} := u] = C$,
possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$
of $t$.

@@

 \subsubsection{Syntactic model}
 The syntactic model is obtained by packing the untyped syntax with the typing
 judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\GG \GG\vdash
 \\
 & \syn\Ty \,(\GG,\wf\GG)&& := \sum_A \GG\vdash A
 \\
 & \syn\Tm \,(\GG,\wf\GG)(A,\wf A)&& := \sum_t \GG\vdash t \in A
 \\
 & \syn\Sub \,(\GG,\wf\GG)(\GD,\wf \GD)&& := \sum_\sigma \GG\vdash \sigma \Rightarrow \GD
\end{alignat*}
The other fields are given straightforwardly.  Regarding the equations, it is
enough to prove them only for the untyped syntactic part: as we argued in
Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are
automatically equal.

@@

Given a model $M$ of the QIIT, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

For a type $A$ in a context $\GG$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\GG}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\,\,\GG$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\GG\sim\model\GG$ and $\rec\,\,\GG=\model\GG$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\GG$ which is
related to $\GG$. Then, the type of the relation for types is
\[
  (\GG : \syn\Con)\ra (A : \syn\Ty\,\syn\GG) \ra
  (\model\GG : \model\Con)
  \ra
  (\GG \sim \model\GG)
  \ra
  (\model{A}:\model\Ty\,\model\GG)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
We want to avoid using such recursive-recursive
definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
$\GG\sim\model\GG$ from the list of arguments.  We proceed similarly for
terms and substitutions. Actually, this removal is not without harm. For
example, consider relating the empty substitution
$\GG\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
$\model\sigma : \model\Sub\,\model\GG\,\model\GD$. We would like to assert
that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
this is not possible because typechecking requires that $\model\GD$ is the
empty semantic context. This is precisely what was ensured by the hypothesis
$\syn\cdot \sim \model\GD$ we removed.  Our way out here is to state that $\model\sigma$
is related to the empty substitution if the target semantic context $\model\GD$
is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.
